---
title: "Bounding treatment effects when subjects go missing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bounding treatment effects when subjects go missing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
options(digits = 4)

# vayr and estimatr are in Suggests, so the chunks that use them are guarded on
# both being installed and on vayr being new enough to have
# impute_extreme_values(). Everything else runs either way.
has_suggests <-
  requireNamespace("vayr", quietly = TRUE) &&
  requireNamespace("estimatr", quietly = TRUE) &&
  utils::packageVersion("vayr") >= "1.1.0"
```

```{r setup}
library(attrition)
library(ggplot2)
library(dplyr)
library(purrr)
```

```{r, eval = has_suggests}
library(vayr)
library(estimatr)
```

This vignette works through a two-wave survey experiment in which 536 of 1,980 subjects did not report an outcome. The example is the replication study from Coppock, Gerber, Green, and Kern (2017), which ships with the package as `levendusky_replication`.

We work through the extreme value bounds without double sampling and how the extreme value bounds change with the second round data collection. In short, a follow-up on 100 of the 536 nonrespondents shrinks the identification region by a factor of 3.6, from 3.25 points wide to 0.91 on an outcome that ranges from 0 to 6.

## The experiment

`levendusky_replication` dataset is derived from a two-wave survey experiment run on Mechanical Turk, replicating Levendusky and Malhotra (2016). Subjects read a news article describing the electorate as sharply divided (the polarized condition) or as focused on common ground (the moderate condition). The outcome, "perceived polarization," was measured immediately and again ten days later. (The experiment ran a third group who read nothing on the topic, which the paper does not analyze and which is not included here, so the 1,980 rows represent only the polarized-versus-moderate contrast.)

Here's a description of the dataset:

| Column | Description |
|---|---|
| `X_party_id` | Party identification in three categories, used below for poststratification. |
| `Z_condition` | The condition as assigned: `Moderate` or `Polarized`. |
| `Z` | Polarized (1) versus moderate (0). |
| `Y_polarization_w2` | Perceived polarization at Wave 2, scored 0 to 6, with some missingness. |
| `R1` | Answered the second wave on the first attempt. |
| `Attempt` | Drawn into the follow-up sample and offered the larger incentive. |
| `R2` | Answered the follow-up attempt. |


```{r}
levendusky_replication |>
  count(Z_condition, R1)
```

Of 1,980 subjects, 1,444 answered the second-wave survey and 536 did not. A naive difference in means among those 1,444 respondents puts the effect of the polarized article at 0.126 points (standard error 0.066), which is significant at the 10 percent level and not at the 5 percent level.

```{r, eval = has_suggests}
naive_fit <-
  difference_in_means(formula = Y_polarization_w2 ~ Z,
                      data = filter(levendusky_replication, R1 == 1))
tidy(naive_fit)
```

This estimate is naive in the sense that it only estimates the average treatment effect if wave 2 attrition is unrelated to potential outcomes.

## Extreme value bounds

If we are unwilling to make that assumption, we can estimate extreme value bounds (Manski, 1990). The outcome runs from 0 to 6. Filling in every missing outcome in the treatment group with 0 and every missing outcome in the control group with 6 gives the lowest average effect the data can support. Doing the opposite yields the highest possible estimate. `estimator_ev` implements this approach and reports an Imbens-Manski confidence interval around the resulting identification region.

```{r}
estimator_ev(Y = Y_polarization_w2,
             Z = Z,
             R = R1,
             minY = 0,
             maxY = 6,
             data = levendusky_replication)
```

The bounds range from -1.54 and 1.71. The bounds are nearly useless: they are wide enough to contain a large negative effect, a large positive effect, and everything in between. The confidence interval is still wider, running from -1.67 to 1.84.

We can visualize the bounds using an approach from the `vayr` package. `R1` rather than the outcome column decides who is missing, and the two disagree for sixteen subjects who started the Wave 2 survey and abandoned it partway: they have an outcome because the outcome averages the policy items a subject did answer, and `R1 == 0` because they never finished. Blanking those outcomes puts the picture and the estimator on the same 536 missing values.

```{r, eval = has_suggests}
bounded <-
  levendusky_replication |>
  mutate(Y_polarization_w2 = if_else(R1 == 1, Y_polarization_w2, NA_real_)) |>
  impute_extreme_values(outcome = "Y_polarization_w2",
                        assignment = "Z",
                        range = c(0, 6))

bound_means <-
  bounded |>
  group_by(Z_condition, scenario) |>
  summarise(Y_polarization_w2 = mean(Y_polarization_w2), .groups = "drop")
```

```{r, eval = has_suggests, fig.alt = "Two panels of the same 1,980 subjects. In the lower bound panel the imputed outcomes sit at 0 for the polarized group and at 6 for the moderate group; the upper bound panel reverses them."}
# Labelled in the left panel only. The colours carry over to the right one, and
# a second copy of the same two words would just be more ink.
label_df <-
  bounded |>
  distinct(scenario, imputed) |>
  filter(scenario == "Lower bound") |>
  mutate(
    Z_condition = "Moderate",
    Y_polarization_w2 = if_else(imputed == "Outcome imputed", 6.4, 4.9),
    label = if_else(imputed == "Outcome imputed", "imputed", "reported")
  )

ggplot(data = bounded,
       mapping = aes(x = Z_condition, y = Y_polarization_w2)) +
  geom_point(mapping = aes(colour = imputed, shape = imputed),
             position = position_jitter(width = 0.25, height = 0),
             alpha = 0.5, stroke = 0) +
  geom_point(data = bound_means, size = 3) +
  geom_text(data = label_df,
            mapping = aes(label = label, colour = imputed),
            hjust = 0, nudge_x = 0.15, size = 3.2, show.legend = FALSE) +
  facet_wrap(facets = ~ scenario) +
  scale_colour_manual(values = c("#205C8A", "#C67800")) +
  scale_y_continuous(breaks = 0:6) +
  labs(x = NULL, y = "Perceived polarization (0 to 6)") +
  theme_minimal() +
  theme(legend.position = "none")
```

The orange points are the 536 imputations. In the left panel they sit at the bottom of the scale for the polarized group and at the top for the moderate group, which is the least favorable arrangement the data admit. The right panel reverses them. The black points are the arm means either way, and the gap between them within a panel is that panel's bound.

Those gaps are the estimates, not an illustration of them. A difference in means run inside each panel recovers both bound estimates exactly, because filling in the missing outcomes and averaging is all the estimator does to reach its point estimates.

```{r, eval = has_suggests}
bounded |>
  group_by(scenario) |>
  reframe(tidy(lm_robust(formula = Y_polarization_w2 ~ Z))) |>
  filter(term == "Z") |>
  select(scenario, difference_in_means = estimate)
```

The confidence interval is not drawn twice the same way. The tempting move is to put an ordinary interval around each panel's difference in means and take the outermost endpoints, which is wider than it needs to be. The Imbens-Manski interval covers the true effect with probability 0.95 rather than covering the whole identified set, and it uses the fact that the effect cannot sit at both ends at once.

```{r, eval = has_suggests}
ev <- estimator_ev(Y = Y_polarization_w2,
                   Z = Z,
                   R = R1,
                   minY = 0,
                   maxY = 6,
                   data = levendusky_replication)

bind_rows(
  bounded |>
    group_by(scenario) |>
    reframe(tidy(lm_robust(formula = Y_polarization_w2 ~ Z))) |>
    filter(term == "Z") |>
    summarise(interval = "Stacking the two panels",
              lower = min(conf.low),
              upper = max(conf.high)),
  tidy(ev) |>
    filter(term == "bounds") |>
    transmute(interval = "Imbens-Manski", lower = conf.low, upper = conf.high)
) |>
  mutate(width = upper - lower)
```

## Double sampling

The design is Neyman's (1938), turned on survey nonresponse by Hansen and Hurwitz (1946). After the initial round of data collection, a random sample of the nonrespondents was drawn and pursued with more effort. In this study, 50 nonrespondents were drawn at random from each condition and offered \$4.00 instead of the original \$1.00. Of those 100 subjects, 72 answered.

```{r}
levendusky_replication |>
  count(Z_condition, Attempt, R2)
```

Because the follow-up sample was drawn at random from the nonrespondents, the outcomes it recovers estimate the mean outcome among *all* nonrespondents. Only the subjects who refused twice (28 subjects) still need worst-case imputation.

```{r}
estimator_ds(Y = Y_polarization_w2,
             Z = Z,
             R1 = R1,
             Attempt = Attempt,
             R2 = R2,
             minY = 0,
             maxY = 6,
             data = levendusky_replication)
```

The identification region is now -0.34 to 0.57, against -1.54 to 1.71 before. It has shrunk by a factor of 3.6, and the confidence interval has narrowed from 3.50 points wide to 1.27, on a follow-up of 100 subjects out of 536.

The interval still contains zero, so the study does not establish that the polarized article changed perceived polarization. What it does establish is a bound on how large any such effect could be, and that bound is now tight enough to be substantively informative.

## Poststratification

A discrete covariate that predicts the outcome can increase the precision further. The bounds are estimated separately inside each of its categories and then averaged using the share of the sample falling in each, which is poststratification in the sense of Miratrix, Sekhon, and Yu (2013).

```{r}
estimator_ds(Y = Y_polarization_w2,
             Z = Z,
             R1 = R1,
             Attempt = Attempt,
             R2 = R2,
             strata = X_party_id,
             minY = 0,
             maxY = 6,
             data = levendusky_replication)
```

The upper bound estimate falls from 0.57 to 0.53 and the confidence interval from 1.27 to 1.23 points wide. The gain is real but modest, which is what to expect when the covariate is only moderately prognostic.

These three sets of numbers are Table 3 of [the published paper](https://doi.org/10.1017/pan.2016.6).

## Sensitivity analysis

Worst-case bounds assume nothing about the outcomes the 28 subjects who refused twice would have reported. Ignorability assumes that their outcomes look like those of the follow-up respondents. `estimator_ds_sens` interpolates between the two assumptions. `delta` is the fraction of follow-up nonrespondents whose outcomes are left unmodeled, and the remaining 1 - `delta` are treated as ignorable.

At `delta = 1` the estimator reproduces the double-sampling bounds above, -0.34 to 0.57, since refusing to model any of the follow-up nonrespondents is exactly what `estimator_ds` does. At `delta = 0` it returns a point estimate.

```{r}
estimator_ds_sens(Y = Y_polarization_w2,
                  Z = Z,
                  R1 = R1,
                  Attempt = Attempt,
                  R2 = R2,
                  delta = 0.5,
                  minY = 0,
                  maxY = 6,
                  data = levendusky_replication)
```

`sensitivity_ds` sweeps `delta` from 0 to 1 and looks for delta\*, the smallest value at which the confidence interval starts to include zero. A delta\* near zero means the finding rests on assuming away nearly all of the missingness; a delta\* near one means it survives almost any amount.

```{r, fig.alt = "Identification regions and confidence intervals as a function of the sensitivity parameter delta"}
sens <-
  sensitivity_ds(Y = Y_polarization_w2,
                 Z = Z,
                 R1 = R1,
                 Attempt = Attempt,
                 R2 = R2,
                 minY = 0,
                 maxY = 6,
                 alpha = 0.10,
                 data = levendusky_replication)

sens
sens$sensitivity_plot
```

The plot reads left to right, from ignorability at `delta = 0` to the worst case at `delta = 1`. The two lines are the lower and upper bound estimates, which coincide at the left edge, where the estimator returns a point, and separate as more of the follow-up nonrespondents are left unmodeled. The shaded band is the confidence interval around them, and delta\* is marked at the point where that band first reaches the dashed line at zero.

At the 10 percent level, delta\* is 0.07. The naive result is fragile: allowing ignorability to fail for 7 percent of the follow-up nonrespondents is enough to erase it. At the 5 percent level there is no delta\* at all, because the interval around the naive estimate already includes zero, and `delta_star` is `NA`.

```{r}
sensitivity_ds(Y = Y_polarization_w2,
               Z = Z,
               R1 = R1,
               Attempt = Attempt,
               R2 = R2,
               minY = 0,
               maxY = 6,
               alpha = 0.05,
               data = levendusky_replication)$delta_star
```

## Trimming bounds

Trimming bounds (Lee, 2009) bracket the effect among subjects who would respond in either condition, the so-called "always reporters." The classic version runs on monotonicity: treatment moves response in one direction only. If treatment can only make subjects more likely to respond, never less, then every control respondent is an always reporter, the extra respondents treatment produced are a known fraction of the treated respondents, and trimming that fraction off either tail of the treated distribution bounds the effect among the always reporters.

Monotonicity has a direction, and the estimator has to be told which one it is. The two directions are not two ways of writing the same assumption. Each names a different group as the always reporters and trims the other, so they give different bounds on the same data.

```{r}
levendusky_replication |>
  group_by(Z_condition) |>
  summarise(response_rate = mean(R1))
```

The polarized response rate is 0.724 compared with 0.735 in the control group. The default direction, `monotonicity = "treatment_increases_response"`, is the one Lee states, and it implies a trimming proportion that these response rates make negative. The estimator returns `NA` for every quantity rather than a number built on an assumption the data contradict, and says which direction they leave open.

```{r}
estimator_trim(Y = Y_polarization_w2,
               Z = Z,
               R = R1,
               data = levendusky_replication)[c("estimate_lower", "estimate_upper")]
```

Assuming instead that the polarized article can only have lowered response puts the control group in the trimmed role, and the estimator runs.

```{r}
estimator_trim(Y = Y_polarization_w2,
               Z = Z,
               R = R1,
               monotonicity = "treatment_decreases_response",
               data = levendusky_replication)[c("estimate_lower", "estimate_upper")]
```

The bounds are 0.08 to 0.16, far narrower than anything above, because only 1.5 percent of control respondents get trimmed. Narrow bounds bought this way are worth reading carefully. The response rates did not choose the direction and cannot: a gap of 0.011 between two rates estimated on about a thousand subjects each is well inside sampling noise, and even a large gap would be a consequence of the assumption rather than evidence for it. Monotonicity is a claim about how the article affected the decision to answer, and it holds or fails whatever the two rates happen to be.

### Dropping monotonicity

`monotonicity = "none"` refuses the direction as well, and assumes only that treatment was randomly assigned. What survives is a lower bound on how many always reporters there can be: at least a share 1 - f0 - f1 of the sample, where fz is the missingness rate in arm z, since two events covering 1 - f0 and 1 - f1 of the sample must overlap by at least that much. Each arm is then trimmed by the largest share of its own respondents that could fail to be always reporters, f0/(1 - f1) of the treatment group and f1/(1 - f0) of the control group. Those are the sharp bounds of Imai (2008, Proposition 1), which build on Zhang and Rubin (2003) and Horowitz and Manski (1995). Since both arms are trimmed by the same rule, this version has no direction to set, and it exists only while f0 + f1 is below 1.

The assumption and the design are separate choices, so there are four estimators here rather than two, and `estimator_trim` reaches all four.

```{r}
cells <- list(
  "single sample, monotonicity" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R = R1,
                   monotonicity = "treatment_decreases_response",
                   se = "none", data = levendusky_replication),
  "single sample, no monotonicity" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R = R1,
                   monotonicity = "none",
                   se = "none", data = levendusky_replication),
  "double sampling, monotonicity" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                   monotonicity = "treatment_decreases_response",
                   se = "none", data = levendusky_replication),
  "double sampling, no monotonicity" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                   monotonicity = "none",
                   se = "none", data = levendusky_replication)
)

tibble(
  cell = names(cells),
  lower = map_dbl(cells, "estimate_lower"),
  upper = map_dbl(cells, "estimate_upper"),
  width = upper - lower
)
```

The two narrowing devices are doing different work. Double sampling narrows by recovering outcomes: it cuts the share of subjects who never reported from 27 percent to 1.4 percent, which is why the assumption-free bounds go from 2.98 points wide to 0.84. Monotonicity narrows by assumption, and it narrows most where there is most left to assume about, which is the single sample. Read the rows against each other with one caveat: the always reporters are not the same people in every row, because who counts as one is part of what the assumption fixes.

### Standard errors

Standard errors come two ways. `se = "analytic"` uses the closed-form variance in Lee (2009, Proposition 3), which is derived for one design and one assumption: a single unweighted sample with one group trimmed. It is the default, and it is available in exactly that cell. `se = "bootstrap"` resamples units within treatment arm and works everywhere, which is what the other three cells need, and why the call below sets a seed: without one the standard errors move a little from run to run. Asking for analytic standard errors where they do not apply is an error rather than a silent substitution.

```{r}
set.seed(343)
estimator_trim(Y = Y_polarization_w2,
               Z = Z,
               R1 = R1,
               Attempt = Attempt,
               R2 = R2,
               se = "bootstrap",
               sims = 500,
               data = levendusky_replication)[
                 c("estimate_lower", "estimate_upper",
                   "std.error_lower", "std.error_upper")]
```

## Reading the output

Every estimator returns a named numeric vector, and every one returns the same six elements under the same names.

```{r}
out <- estimator_ds(Y = Y_polarization_w2,
                    Z = Z,
                    R1 = R1,
                    Attempt = Attempt,
                    R2 = R2,
                    minY = 0,
                    maxY = 6,
                    data = levendusky_replication)
out
```

`estimate_lower` and `estimate_upper` are the two ends of the identification region, `std.error_lower` and `std.error_upper` are their standard errors, and `conf.low` and `conf.high` are the joint Imbens-Manski interval. Note that the third pair are standard errors rather than variances. The names are broom's, which is the point of them: `tidy()` returns the same six quantities in a data frame, under exactly the same names.

```{r}
tidy(out)
```

Bounds do not have a point estimate, so `estimate` is `NA` on the `bounds` row, which is the vector above laid out across columns. The two rows below split it, one endpoint each, so that `estimate` and `std.error` mean on those rows what broom means by them, and `DeclareDesign::declare_estimator()` can select a single endpoint with `term`.

Each estimator also takes a formula, which is what `declare_estimator()` expects. When the first argument is a formula, the remaining columns are named as strings rather than passed bare, so `R1` becomes `R1 = "R1"`. The interface changed, not the argument.

```{r}
estimator_ds(Y = Y_polarization_w2 ~ Z,
             R1 = "R1",
             Attempt = "Attempt",
             R2 = "R2",
             minY = 0,
             maxY = 6,
             data = levendusky_replication)
```

## What the design costs and what it buys

Double sampling is not free. Someone has to find the nonrespondents and pay them more, and the budget for that effort competes with the budget for a larger initial sample. The trade is usually worth making, because a larger initial sample does nothing about attrition bias while a follow-up sample reduces it directly. In this study the initial-sample confidence interval was 3.50 points wide and no realistic increase in sample size would have narrowed it much, since almost all of that width came from the 536 unknown outcomes rather than from sampling error. A follow-up on 100 of them narrowed it to 1.27, and poststratification to 1.23.

The design decision that remains open is how to split a fixed budget between the initial sample and the follow-up, and it depends on quantities not known until the data are in: the response rate in each round, the outcome variance among respondents and nonrespondents, and the relative cost of a first and second contact. Coppock, Gerber, Green, and Kern (2017) sketch the optimization; the package does not yet implement it.

## All of the estimators at once

Nine estimates have gone past, and they do not all estimate the same thing. Two questions separate them. *Whose* effect is it: everyone who took Wave 1, or only the subjects who would have reported either way, or only the ones who did report? And *what is assumed* to get there?

| Estimator | Estimand | Assumption |
|---|---|---|
| Naive difference in means | Effect among Wave 2 respondents | Attrition is ignorable |
| Extreme value | Effect among all 1,980 | Outcome lies in [0, 6] |
| Extreme value, double sampled | Effect among all 1,980 | The above, plus a random follow-up sample |
| Extreme value, double sampled, poststratified | Effect among all 1,980 | The above; party identification only sharpens the estimate |
| Sensitivity at delta = 0.5 | Effect among all 1,980 | The above, plus ignorability for half the follow-up nonrespondents |
| Trimming, monotonicity | Effect among always reporters | The article never raised the chance of responding |
| Trimming, monotonicity, double sampled | Effect among always reporters | The above, plus a random follow-up sample |
| Trimming, no monotonicity | Effect among always reporters | Random assignment only |
| Trimming, no monotonicity, double sampled | Effect among always reporters | Random assignment plus a random follow-up sample |

The figure below shortens "double sampled" to DS.

```{r}
set.seed(343)

fits <- list(
  "Extreme value" =
    estimator_ev(Y = Y_polarization_w2, Z = Z, R = R1,
                 minY = 0, maxY = 6, data = levendusky_replication),
  "Extreme value, double sampled" =
    estimator_ds(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                 minY = 0, maxY = 6, data = levendusky_replication),
  "Extreme value, DS, poststratified" =
    estimator_ds(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                 strata = X_party_id, minY = 0, maxY = 6, data = levendusky_replication),
  "Sensitivity at delta = 0.5" =
    estimator_ds_sens(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                      delta = 0.5, minY = 0, maxY = 6, data = levendusky_replication),
  "Trimming, monotonicity" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R = R1,
                   monotonicity = "treatment_decreases_response",
                   se = "bootstrap", sims = 500, data = levendusky_replication),
  "Trimming, monotonicity, DS" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                   monotonicity = "treatment_decreases_response",
                   se = "bootstrap", sims = 500, data = levendusky_replication),
  "Trimming, no monotonicity" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R = R1, monotonicity = "none",
                   se = "bootstrap", sims = 500, data = levendusky_replication),
  "Trimming, no monotonicity, DS" =
    estimator_trim(Y = Y_polarization_w2, Z = Z, R1 = R1, Attempt = Attempt, R2 = R2,
                   monotonicity = "none",
                   se = "bootstrap", sims = 500, data = levendusky_replication)
)

estimands <- tribble(
  ~estimator,                          ~estimand,
  "Naive difference in means",         "Wave 2\nrespondents",
  "Extreme value",                     "All 1,980\nsubjects",
  "Extreme value, double sampled",     "All 1,980\nsubjects",
  "Extreme value, DS, poststratified", "All 1,980\nsubjects",
  "Sensitivity at delta = 0.5",        "All 1,980\nsubjects",
  "Trimming, monotonicity",            "Always\nreporters",
  "Trimming, monotonicity, DS",        "Always\nreporters",
  "Trimming, no monotonicity",         "Always\nreporters",
  "Trimming, no monotonicity, DS",     "Always\nreporters"
)

# The naive estimator claims a point rather than a region, which the plot shows as
# an identification region of zero width
naive <- lm(Y_polarization_w2 ~ Z, data = filter(levendusky_replication, R1 == 1))

gg_df <-
  fits |>
  map(\(fit) filter(tidy(fit), term == "bounds")) |>
  list_rbind(names_to = "estimator") |>
  select(estimator, estimate_lower, estimate_upper, conf.low, conf.high) |>
  add_row(estimator = "Naive difference in means",
          estimate_lower = coef(naive)[["Z"]],
          estimate_upper = coef(naive)[["Z"]],
          conf.low = confint(naive)["Z", 1],
          conf.high = confint(naive)["Z", 2],
          .before = 1) |>
  left_join(estimands, by = "estimator") |>
  mutate(
    estimator = factor(estimator, levels = rev(estimands$estimator)),
    estimand = factor(estimand, levels = unique(estimands$estimand))
  )
```

```{r, fig.height = 5, fig.alt = "Identification regions and confidence intervals for nine estimators, grouped by the population each one is about."}
ggplot(data = gg_df,
       mapping = aes(y = estimator)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey65") +
  geom_linerange(mapping = aes(xmin = conf.low, xmax = conf.high),
                 linewidth = 0.5, colour = "#9AA5AE") +
  geom_linerange(mapping = aes(xmin = estimate_lower, xmax = estimate_upper),
                 linewidth = 2.4, colour = "#205C8A") +
  geom_point(data = filter(gg_df, estimate_lower == estimate_upper),
             mapping = aes(x = estimate_lower), size = 2.2, colour = "#205C8A") +
  facet_grid(rows = vars(estimand), scales = "free_y", space = "free_y", switch = "y") +
  labs(x = "Effect on perceived polarization",
       y = NULL,
       subtitle = "Thick: identification region\nThin: 95 percent confidence interval") +
  theme_minimal(base_size = 10) +
  theme(strip.placement = "outside",
        strip.text.y.left = element_text(angle = 0, hjust = 0),
        panel.grid.major.y = element_blank(),
        plot.subtitle = element_text(size = 8.5, colour = "grey30"))
```

Read down the panels rather than across them. Inside a panel the estimators are comparable, and the pattern is the one the paper is about: the assumption-free region is uselessly wide, the follow-up shrinks it by a factor of 3.6 at a cost of chasing 100 people, and poststratification tightens the interval a little more. The sensitivity row sits inside the double-sampling row because it assumes away half of what that row leaves open.

Across panels the comparison is not like for like. The naive point sits in a panel of its own because it describes the 1,444 subjects who answered, a group selected after treatment, and it is a point rather than a region only because ignorability is assumed rather than shown. The trimming rows sit in a third panel because they describe the always reporters, a subgroup whose membership shifts with the assumption used to bound it, which is why a narrow trimming interval and a wide extreme value interval are not evidence about each other.

One detail visible only in the console: the bootstrap for the double-sampled trimming row discards the resamples in which the response rates cross over and monotonicity fails in the assumed direction. A gap of 0.011 between the two rates is small enough that this happens often, and the standard error is computed from the replicates that survived. The direction is an assumption these data are in no position to support.

## References

Coppock, Alexander, Alan S. Gerber, Donald P. Green, and Holger L. Kern (2017). Combining Double Sampling and Bounds to Address Nonignorable Missing Outcomes in Randomized Experiments. *Political Analysis* 25(2):188-206. <https://doi.org/10.1017/pan.2016.6>

Hansen, Morris H., and William N. Hurwitz (1946). The Problem of Non-Response in Sample Surveys. *Journal of the American Statistical Association* 41(236):517-529. <https://doi.org/10.1080/01621459.1946.10501894>

Imbens, Guido W., and Charles F. Manski (2004). Confidence Intervals for Partially Identified Parameters. *Econometrica* 72(6):1845-1857. <https://doi.org/10.1111/j.1468-0262.2004.00555.x>

Lee, David S. (2009). Training, Wages, and Sample Selection: Estimating Sharp Bounds on Treatment Effects. *Review of Economic Studies* 76(3):1071-1102. <https://doi.org/10.1111/j.1467-937X.2009.00536.x>

Levendusky, Matthew, and Neil Malhotra (2016). Does Media Coverage of Partisan Polarization Affect Political Attitudes? *Political Communication* 33(2):283-301. <https://doi.org/10.1080/10584609.2015.1038455>

Manski, Charles F. (1990). Nonparametric Bounds on Treatment Effects. *American Economic Review Papers and Proceedings* 80(2):319-323.

Miratrix, Luke W., Jasjeet S. Sekhon, and Bin Yu (2013). Adjusting Treatment Effect Estimates by Post-Stratification in Randomized Experiments. *Journal of the Royal Statistical Society, Series B* 75(2):369-396. <https://doi.org/10.1111/j.1467-9868.2012.01048.x>

Neyman, Jerzy (1938). Contribution to the Theory of Sampling Human Populations. *Journal of the American Statistical Association* 33(201):101-116. <https://doi.org/10.1080/01621459.1938.10503378>

Horowitz, Joel L., and Charles F. Manski (1995). Identification and Robustness with Contaminated and Corrupted Data. *Econometrica* 63(2):281-302. <https://doi.org/10.2307/2951627>

Imai, Kosuke (2008). Sharp Bounds on the Causal Effects in Randomized Experiments with "Truncation-by-Death". *Statistics & Probability Letters* 78(2):144-149. <https://doi.org/10.1016/j.spl.2007.05.015>

Zhang, Junni L., and Donald B. Rubin (2003). Estimation of Causal Effects via Principal Stratification When Some Outcomes are Truncated by "Death". *Journal of Educational and Behavioral Statistics* 28(4):353-368. <https://doi.org/10.3102/10769986028004353>
