---
title: "Bounding treatment effects when subjects go missing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bounding treatment effects when subjects go missing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
options(digits = 4)
```

```{r setup}
library(attrition)
```

This vignette works through a two-wave survey experiment in which 536 of 1,980 subjects never answered the second wave. Everything runs on the replication study from Coppock, Gerber, Green, and Kern (2017), which ships with the package as `levendusky`.

The situation is an ordinary one. You fielded an experiment, the outcome is missing for a substantial minority of subjects, and nothing in the data tells you whether they went missing for reasons connected to the outcomes they would have reported. Rather than settle that question by assumption, the estimators here report the range of average treatment effects consistent with what you did observe. The double-sampling design then shrinks that range, by recovering outcomes from a random sample of the nonrespondents and pursuing them harder than the first attempt.

We take the estimators in the order you would meet the problem: what the data look like, what can be said with no follow-up at all, what a follow-up buys, and how much of the remaining assumption a result survives. The short version is that chasing 100 of the 536 nonrespondents shrinks the identification region by a factor of 3.6, from 3.25 points wide to 0.91 on an outcome scored from 0 to 6.

## The experiment

`levendusky` holds a two-wave survey experiment run on Mechanical Turk, replicating Levendusky and Malhotra (2016). Subjects read a news article describing the electorate as sharply divided (the polarized condition) or as focused on common ground (the moderate condition). A third group read nothing on the topic and is not analyzed. The outcome, perceived polarization, was measured immediately and again ten days later.

Seven of its twelve columns do the work below. `?levendusky` documents them all.

| Column | What it holds |
|---|---|
| `L_dif_w2` | Perceived polarization at Wave 2, scored 0 to 6. The outcome, and the only column with missing values. |
| `Z_lev` | The condition as assigned: `Placebo`, `Moderate`, or `Polarized`. |
| `Z1` | Polarized (1) versus moderate (0), and `NA` in the placebo condition. The contrast the paper analyzes. |
| `R1` | Answered the second wave on the first attempt. |
| `Attempt` | Drawn into the follow-up sample and offered the larger incentive. |
| `R2` | Answered the follow-up attempt. |
| `pid_3_recoded` | Party identification in three categories, used below for poststratification. |

Every analysis in the paper is the polarized-versus-moderate contrast, so the placebo group goes first. Dropping it is what `!is.na(Z1)` does: `Z1` is defined only for the two conditions being compared, so filtering on its missingness removes the placebo group and nothing else. `droplevels()` then clears the emptied factor level, which would otherwise print as a row of zeroes in every table that follows.

```{r}
dat <- droplevels(subset(levendusky, !is.na(Z1)))
with(dat, table(Z_lev, R1))
```

Of 1,980 subjects, 1,444 answered the second-wave survey and 536 did not. A naive difference in means among those 1,444 respondents puts the effect of the polarized article at 0.126 points (standard error 0.066), which is significant at the 10 percent level and not at the 5 percent level.

```{r}
summary(lm(L_dif_w2 ~ Z1, data = subset(dat, R1 == 1)))$coefficients
```

The subset matters. Restricting to `R1 == 1` is what makes the estimate naive: it is the analysis you would run if the double sampling had never happened.

That estimate is consistent only if the 536 missing subjects are missing for reasons unrelated to their outcomes. Nothing in the data tells us whether they are.

## Worst-case bounds

Suppose we refuse the assumption entirely. The outcome runs from 0 to 6, so we know something about the missing values even without knowing what they are: whatever they were, they were between 0 and 6. Fill in every missing outcome in the treatment group with 0 and every missing outcome in the control group with 6, and you get the lowest average effect the data can support. Reverse the fills and you get the highest.

Those two numbers do all the work below, so it is worth being clear about where they come from. They are properties of the measurement instrument rather than of this sample. Perceived polarization is an average of absolute differences between two ratings of the same policy question, and such a difference can be neither negative nor wider than the rating scale, which fixes the limits at 0 and 6 whatever the data happen to contain. Here the data happen to contain both.

```{r}
range(dat$L_dif_w2, na.rm = TRUE)
```

`estimator_ev` does the filling, and reports an Imbens-Manski confidence interval around the resulting identification region.

```{r}
estimator_ev(L_dif_w2, Z1, R1, minY = 0, maxY = 6, data = dat)
```

The effect lies between -1.54 and 1.71. The bounds are honest and nearly useless: they are wide enough to contain a large negative effect, a large positive effect, and everything in between. Notice how little the sampling uncertainty contributes. The confidence interval runs from -1.67 to 1.84, barely wider than the bounds themselves. The problem here is not that the sample is small. The problem is that 536 outcomes are unknown, and no sample size fixes that.

The companion vignette, `vignette("drawing-the-bounds")`, draws the imputation this estimator averages over, on a smaller simulated experiment where the picture fits on one page.

## Double sampling

Now the design. After the initial round of data collection, draw a random sample of the nonrespondents and pursue them with more effort than you spent the first time. In this study, 50 nonrespondents were drawn at random from each condition and offered \$4.00 instead of the original \$1.00. Of those 100 subjects, 72 answered.

```{r}
with(dat, table(Z_lev, Attempt, R2))
```

The 72 recovered outcomes do more work than their number suggests. Because the follow-up sample was drawn at random from the nonrespondents, the outcomes it recovers estimate the mean outcome among *all* nonrespondents. Only the subjects who refused twice, and there are 28 of them, still need worst-case imputation.

```{r}
estimator_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6, data = dat)
```

The identification region is now -0.34 to 0.57, against -1.54 to 1.71 before. It has shrunk by a factor of 3.6, and the confidence interval has narrowed from 3.50 points wide to 1.27. Chasing 100 subjects out of 536 bought all of that.

To be sure, the interval still contains zero, so the study does not establish that the polarized article changed perceived polarization. What it does establish is a bound on how large any such effect could be, and that bound is now tight enough to be substantively informative.

## Poststratification

If you measured a discrete covariate that predicts the outcome, use it. Estimate the bounds separately inside each of its categories, then average those bounds using the share of the sample falling in each. The estimand does not change, so nothing is assumed away, but the estimate of it gets more precise.

The precision comes from where the category shares are estimated. Write `B` for the covariate and Pr(B = k) for the share of the whole sample in category k. Poststratification weights by that whole-sample share, which every subject helps to pin down. The unadjusted estimator effectively weights by Pr(B = k | Z = z), the share within one arm at a time, and each arm is half the sample. By the law of total variance, trading the arm-specific shares for the whole-sample ones is guaranteed not to raise the asymptotic variance. Here the covariate is three-category party identification.

```{r}
estimator_ds(L_dif_w2, Z1, R1, Attempt, R2, strata = pid_3_recoded,
             minY = 0, maxY = 6, data = dat)
```

The upper bound estimate falls from 0.57 to 0.53 and the confidence interval from 1.27 to 1.23 points wide. The gain is real but modest, which is what to expect when the covariate is only moderately prognostic. Bear in mind what moved and what did not: the estimand is the same identification region in both rows, and poststratification only estimates it better.

These three sets of numbers are Table 3 of [the published paper](https://doi.org/10.1017/pan.2016.6).

## Sensitivity analysis

Worst-case bounds assume nothing about the 28 subjects who refused twice. Ignorability assumes everything: that their outcomes look like those of the follow-up respondents. Neither extreme is a natural place to stand, and `estimator_ds_sens` interpolates between them. Set `delta` to the fraction of follow-up nonrespondents whose outcomes you are unwilling to model, and the remaining 1 - `delta` are treated as ignorable.

At `delta = 1` the estimator reproduces the double-sampling bounds above, -0.34 to 0.57, since refusing to model any of the follow-up nonrespondents is exactly what `estimator_ds` does. At `delta = 0` it returns a point estimate.

```{r}
estimator_ds_sens(L_dif_w2, Z1, R1, Attempt, R2, delta = 0.5,
                  minY = 0, maxY = 6, data = dat)
```

`sensitivity_ds` sweeps `delta` from 0 to 1 and looks for delta\*, the smallest value at which the confidence interval starts to include zero. A delta\* near zero means the finding rests on assuming away nearly all of the missingness; a delta\* near one means it survives almost any amount.

```{r, fig.alt = "Identification regions and confidence intervals as a function of the sensitivity parameter delta"}
sens <- sensitivity_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6,
                       alpha = 0.10, data = dat)
sens$sensitivity_plot
sens$delta_star
```

The plot reads left to right, from ignorability at `delta = 0` to the worst case at `delta = 1`. The two lines are the lower and upper bound estimates, which coincide at the left edge, where the estimator returns a point, and separate as more of the follow-up nonrespondents are left unmodeled. The shaded band is the confidence interval around them, and delta\* is marked at the point where that band first reaches the dashed line at zero.

At the 10 percent level, delta\* is 0.07. The naive result is fragile: allowing ignorability to fail for 7 percent of the follow-up nonrespondents is enough to erase it. At the 5 percent level there is no delta\* at all, because the interval around the naive estimate already includes zero, and `delta_star` is `NA` to say so.

```{r}
sensitivity_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6,
               alpha = 0.05, data = dat)$delta_star
```

## Trimming bounds

Everything so far leans on the outcome having known lower and upper limits. When it does not, or when the limits are so far apart that worst-case bounds say nothing, Lee (2009) offers a different trade: assume instead that treatment moves selection in one direction only. If treatment can only make subjects more likely to respond, never less, then the extra respondents it produces are a known fraction of the treated respondents, and trimming that fraction off either tail of the treated distribution bounds the effect among subjects who would have responded either way.

Monotonicity is an assumption about the world, not about the measurement scale, so it is not nested with the bounded-support assumption. Neither dominates.

Monotonicity is checkable in one direction, at least: it implies the treated respond at least as often as the controls. In this experiment they do not.

```{r}
with(dat, tapply(R1, Z1, mean))
```

The treated responded at 0.724 and the controls at 0.735. The gap is small and could easily be noise, but the trimming proportion it implies is negative, so the assumption fails in the direction the estimator needs and `estimator_trim` returns `NA` rather than a number built on it.

```{r}
estimator_trim(L_dif_w2, Z1, R = R1, data = dat)[c("estimate_lower", "estimate_upper")]
```

Run on the double-sampled data, the same function returns numbers. The reason is that it is not the same estimator: which arguments you supply is the switch between two of them. `R` asks for Lee's single-stage estimator, which trims one group and needs monotonicity to know which. `R1`, `Attempt` and `R2` ask for the double-sampling version, which trims *both* groups, by the share still missing in the other, and carries the follow-up sampling weights. Monotonicity is not among its assumptions, so there is no monotonicity condition left to fail.

```{r}
set.seed(343)
estimator_trim(L_dif_w2, Z1, R1 = R1, Attempt = Attempt, R2 = R2,
               se = "bootstrap", sims = 500, data = dat)[
                 c("estimate_lower", "estimate_upper",
                   "std.error_lower", "std.error_upper")]
```

Standard errors come two ways. `se = "analytic"`, the default, uses the closed-form variance in Lee (2009, Proposition 3), which covers the single-stage case only. `se = "bootstrap"` resamples units within treatment arm and works for both, which is why the double-sampling call above asks for it, and why it also sets a seed: without one the standard errors move a little from run to run. Requesting analytic standard errors on the double-sampling path is an error rather than a silent substitution, since Lee's derivation assumes independent sampling and trimming of one group only.

## Reading the output

Every estimator returns a named numeric vector, and every one returns the same six elements under the same names.

```{r}
out <- estimator_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6, data = dat)
out
```

`estimate_lower` and `estimate_upper` are the two ends of the identification region, `std.error_lower` and `std.error_upper` are their standard errors, and `conf.low` and `conf.high` are the joint Imbens-Manski interval. Note that the third pair are standard errors rather than variances. The names are broom's, which is the point of them: `tidy()` returns the same six quantities in a data frame, under exactly the same names.

```{r}
tidy(out)
```

Bounds do not have a point estimate, so `estimate` is `NA` on the `bounds` row, which is the vector above laid out across columns. The two rows below split it, one endpoint each, so that `estimate` and `std.error` mean on those rows what broom means by them, and `DeclareDesign::declare_estimator()` can select a single endpoint with `term`.

Each estimator also takes a formula, which is what `declare_estimator()` expects. When the first argument is a formula, the remaining columns are named as strings rather than passed bare, so `R1` becomes `R1 = "R1"`. The interface changed, not the argument.

```{r}
estimator_ds(L_dif_w2 ~ Z1, R1 = "R1", Attempt = "Attempt", R2 = "R2",
             minY = 0, maxY = 6, data = dat)
```

## What the design costs and what it buys

Double sampling is not free. Someone has to find the nonrespondents and pay them more, and the budget for that effort competes with the budget for a larger initial sample. The trade is usually worth making, because a larger initial sample does nothing about attrition bias while a follow-up sample attacks it directly. In this study the initial-sample confidence interval was 3.50 points wide and no realistic increase in sample size would have narrowed it much, since almost all of that width came from the 536 unknown outcomes rather than from sampling error. Chasing 100 of them narrowed it to 1.23.

The design decision that remains open is how to split a fixed budget between the initial sample and the follow-up, and it depends on quantities you do not know until the data are in: the response rate in each round, the outcome variance among respondents and nonrespondents, and the relative cost of a first and second contact. Coppock, Gerber, Green, and Kern (2017) sketch the optimization; the package does not yet implement it.

## References

Coppock, Alexander, Alan S. Gerber, Donald P. Green, and Holger L. Kern (2017). Combining Double Sampling and Bounds to Address Nonignorable Missing Outcomes in Randomized Experiments. *Political Analysis* 25(2):188-206. <https://doi.org/10.1017/pan.2016.6>

Imbens, Guido W., and Charles F. Manski (2004). Confidence Intervals for Partially Identified Parameters. *Econometrica* 72(6):1845-1857. <https://doi.org/10.1111/j.1468-0262.2004.00555.x>

Lee, David S. (2009). Training, Wages, and Sample Selection: Estimating Sharp Bounds on Treatment Effects. *Review of Economic Studies* 76(3):1071-1102. <https://doi.org/10.1111/j.1467-937X.2009.00536.x>

Levendusky, Matthew, and Neil Malhotra (2016). Does Media Coverage of Partisan Polarization Affect Political Attitudes? *Political Communication* 33(2):283-301. <https://doi.org/10.1080/10584609.2015.1038455>

Manski, Charles F. (1990). Nonparametric Bounds on Treatment Effects. *American Economic Review Papers and Proceedings* 80(2):319-323.
