---
title: "Bounding treatment effects when subjects go missing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bounding treatment effects when subjects go missing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
options(digits = 4)
```

```{r setup}
library(attrition)
```

Attrition is the problem that covariate adjustment cannot solve. If subjects leave your experiment for reasons related to the outcomes they would have reported, then the respondents you are left with are not a random sample of the subjects you started with, and no regression on the covariates you happened to measure will put that right. The usual fix is an assumption: missingness is ignorable, conditional on treatment and a few covariates. The assumption is untestable, and when it fails it fails silently.

This package takes the other route. Instead of assuming the missing outcomes away, it reports the whole range of average treatment effects consistent with the data, and it implements a research design that makes that range narrow enough to be worth reporting. The design is double sampling: chase a random sample of your nonrespondents, hard.

We work throughout with the replication study from Coppock, Gerber, Green, and Kern (2017), which ships with the package.

## The experiment

`levendusky` holds a two-wave survey experiment run on Mechanical Turk, replicating Levendusky and Malhotra (2016). Subjects read a news article describing the electorate as sharply divided (the polarized condition) or as focused on common ground (the moderate condition). A third group read nothing on the topic and is not analyzed. The outcome, perceived polarization, was measured immediately and again ten days later.

```{r}
dat <- subset(levendusky, !is.na(Z1))
with(dat, table(Z_lev, R1))
```

Of 1,980 subjects, 1,444 answered the second-wave survey and 536 did not. A naive difference in means among those 1,444 respondents puts the effect of the polarized article at 0.126 points (standard error 0.066), which is significant at the 10 percent level and not at the 5 percent level.

```{r}
summary(lm(L_dif_w2 ~ Z1, data = subset(dat, R1 == 1)))$coefficients
```

The subset matters. Restricting to `R1 == 1` is what makes the estimate naive: it is the analysis you would run if the double sampling had never happened.

That estimate is consistent only if the 536 missing subjects are missing for reasons unrelated to their outcomes. Nothing in the data tells us whether they are.

## Worst-case bounds

Suppose we refuse the assumption entirely. The outcome runs from 0 to 6, so we know something about the missing values even without knowing what they are: whatever they were, they were between 0 and 6. Fill in every missing outcome in the treatment group with 0 and every missing outcome in the control group with 6, and you get the lowest average effect the data can support. Reverse the fills and you get the highest.

`estimator_ev` does that, and reports an Imbens-Manski confidence interval around the resulting identification region.

```{r}
estimator_ev(L_dif_w2, Z1, R1, minY = 0, maxY = 6, data = dat)
```

The effect lies between -1.54 and 1.71. The bounds are honest and nearly useless: they are wide enough to contain a large negative effect, a large positive effect, and everything in between. Notice how little the sampling uncertainty contributes. The confidence interval runs from -1.67 to 1.84, barely wider than the bounds themselves. The problem here is not that the sample is small. The problem is that 536 outcomes are unknown, and no sample size fixes that.

## Double sampling

Now the design. After the initial round of data collection, draw a random sample of the nonrespondents and pursue them with more effort than you spent the first time. In this study, 50 nonrespondents were drawn at random from each condition and offered \$4.00 instead of the original \$1.00. Of those 100 subjects, 72 answered.

```{r}
with(dat, table(Z_lev, Attempt, R2))
```

The 72 recovered outcomes do more work than their number suggests. Because the follow-up sample was drawn at random from the nonrespondents, the outcomes it recovers estimate the mean outcome among *all* nonrespondents. Only the subjects who refused twice, and there are 28 of them, still need worst-case imputation.

```{r}
estimator_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6, data = dat)
```

The identification region is now -0.34 to 0.57, against -1.54 to 1.71 before. It has shrunk by a factor of 3.6, and the confidence interval has narrowed from 3.50 points wide to 1.27. Chasing 100 subjects out of 536 bought all of that.

To be sure, the interval still contains zero, so the study does not establish that the polarized article changed perceived polarization. What it does establish is a bound on how large any such effect could be, and that bound is now tight enough to be substantively informative.

## Poststratification

If you measured a discrete covariate that predicts the outcome, use it. Estimating the bounds within each stratum and averaging by stratum share targets the same identification region, so nothing is assumed away, but it estimates that region more precisely. The reason is that the stratum shares Pr(B = k) can be pinned down using the whole sample, while the unadjusted estimator effectively uses Pr(B = k | Z = z), one arm at a time. By the law of total variance the asymptotic variance is guaranteed to be no larger. Here the covariate is three-category party identification.

```{r}
estimator_ds(L_dif_w2, Z1, R1, Attempt, R2, strata = pid_3_recoded,
             minY = 0, maxY = 6, data = dat)
```

The upper bound estimate falls from 0.57 to 0.53 and the confidence interval from 1.27 to 1.23 points wide. The gain is real but modest, which is what to expect when the covariate is only moderately prognostic. Bear in mind what moved and what did not: the estimand is the same identification region in both rows, and poststratification only estimates it better.

These three sets of numbers are Table 3 of the published paper.

## Sensitivity analysis

Worst-case bounds assume nothing about the 28 subjects who refused twice. Ignorability assumes everything: that their outcomes look like those of the follow-up respondents. Neither extreme is a natural place to stand, and `estimator_ds_sens` interpolates between them. Set delta to the fraction of follow-up nonrespondents whose outcomes you are unwilling to model, and the remaining 1 - delta are treated as ignorable.

At delta = 1 the estimator reproduces the worst-case bounds above. At delta = 0 it returns a point estimate.

```{r}
estimator_ds_sens(L_dif_w2, Z1, R1, Attempt, R2, delta = 0.5,
                  minY = 0, maxY = 6, data = dat)
```

`sensitivity_ds` sweeps delta from 0 to 1 and looks for delta\*, the smallest value at which the confidence interval starts to include zero. A delta\* near zero means the finding rests on assuming away nearly all of the missingness; a delta\* near one means it survives almost any amount.

```{r, fig.alt = "Identification regions and confidence intervals as a function of the sensitivity parameter delta"}
sens <- sensitivity_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6,
                       alpha = 0.10, data = dat)
sens$sensitivity_plot
sens$p_star$p
```

At the 10 percent level, delta\* is 0.07. The naive result is fragile: allowing ignorability to fail for 7 percent of the follow-up nonrespondents is enough to erase it. At the 5 percent level there is no delta\* at all, because the interval around the naive estimate already includes zero.

```{r}
sensitivity_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6,
               alpha = 0.05, data = dat)$p_star
```

## Trimming bounds

Everything so far leans on the outcome having known lower and upper limits. When it does not, or when the limits are so far apart that worst-case bounds say nothing, Lee (2009) offers a different trade: assume instead that treatment moves selection in one direction only. If treatment can only make subjects more likely to respond, never less, then the extra respondents it produces are a known fraction of the treated respondents, and trimming that fraction off either tail of the treated distribution bounds the effect among subjects who would have responded either way.

Monotonicity is an assumption about the world, not about the measurement scale, so it is not nested with the bounded-support assumption. Neither dominates.

Monotonicity is checkable in one direction, at least: it implies the treated respond at least as often as the controls. In this experiment they do not.

```{r}
with(dat, tapply(R1, Z1, mean))
```

The treated responded at 0.724 and the controls at 0.735. The gap is small and could easily be noise, but the trimming proportion it implies is negative, so the assumption fails in the direction the estimator needs and `estimator_trim` returns `NA` rather than a number built on it.

```{r}
estimator_trim(L_dif_w2, Z1, R = R1, data = dat)[c("lower_bound", "upper_bound")]
```

Applied to the double-sampled data, where both groups are trimmed and the follow-up respondents carry sampling weights, the bounds exist:

```{r}
estimator_trim(L_dif_w2, Z1, R1 = R1, Attempt = Attempt, R2 = R2,
               se = "bootstrap", sims = 500, data = dat)[c("lower_bound", "upper_bound", "lower_se", "upper_se")]
```

Standard errors come two ways. `se = "analytic"`, the default, uses the closed-form variance in Lee (2009, Proposition 3), which covers the single-stage case only. `se = "bootstrap"` resamples units within treatment arm and works for both, which is why the double-sampling call above asks for it. Requesting analytic standard errors on the double-sampling path is an error rather than a silent substitution, since Lee's derivation assumes independent sampling and trimming of one group only.

## Reading the output

Every estimator returns a named vector, and every one has a `tidy()` method that puts the same information in a data frame.

```{r}
tidy(estimator_ds(L_dif_w2, Z1, R1, Attempt, R2, minY = 0, maxY = 6, data = dat))
```

Bounds do not have a point estimate, so `estimate` is `NA` on the `bounds` row and the identification region appears in `estimate.low` and `estimate.high`. The joint Imbens-Manski interval sits in `conf.low` and `conf.high`. The two rows below carry the bound estimates and their standard errors separately, for the cases where you want one endpoint at a time.

Each estimator also takes a formula, which is what `DeclareDesign::declare_estimator()` expects. Response and attempt indicators are named as strings:

```{r}
estimator_ds(L_dif_w2 ~ Z1, R1 = "R1", Attempt = "Attempt", R2 = "R2",
             minY = 0, maxY = 6, data = dat)
```

## What the design costs and what it buys

Double sampling is not free. Someone has to find the nonrespondents and pay them more, and the budget for that effort competes with the budget for a larger initial sample. The trade is usually worth making, because a larger initial sample does nothing about attrition bias while a follow-up sample attacks it directly. In this study the initial-sample confidence interval was 3.50 points wide and no realistic increase in sample size would have narrowed it much, since almost all of that width came from the 536 unknown outcomes rather than from sampling error. Chasing 100 of them narrowed it to 1.23.

The design decision that remains open is how to split a fixed budget between the initial sample and the follow-up, and it depends on quantities you do not know until the data are in: the response rate in each round, the outcome variance among respondents and nonrespondents, and the relative cost of a first and second contact. Coppock, Gerber, Green, and Kern (2017) sketch the optimization; the package does not yet implement it.

## References

Coppock, Alexander, Alan S. Gerber, Donald P. Green, and Holger L. Kern (2017). Combining Double Sampling and Bounds to Address Nonignorable Missing Outcomes in Randomized Experiments. *Political Analysis* 25(2):188-206.

Imbens, Guido W., and Charles F. Manski (2004). Confidence Intervals for Partially Identified Parameters. *Econometrica* 72(6):1845-1857.

Lee, David S. (2009). Training, Wages, and Sample Selection: Estimating Sharp Bounds on Treatment Effects. *Review of Economic Studies* 76(3):1071-1102.

Levendusky, Matthew, and Neil Malhotra (2016). Does Media Coverage of Partisan Polarization Affect Political Attitudes? *Political Communication* 33(2):283-301.

Manski, Charles F. (1990). Nonparametric Bounds on Treatment Effects. *American Economic Review Papers and Proceedings* 80(2):319-323.
